\section{Homicide}

\begin{enumerate}
    \item Homicide is a neutral term. It is not necessarily a crime.
\end{enumerate}

\subsection{Common Law vs. Model Penal Code}

\begin{tabular}{ | p{5.5cm} | p{5.5cm} |}
\hline
Common Law & Model Penal Code \\ \hline
\begin{enumerate}
    \item \textbf{Murder.}
    \begin{enumerate}
        \item First-degree: premeditated/deliberated.
        \item Second-degree:
        \begin{enumerate}
            \item Intent to kill without premeditation.\footnote{E.g., the 
            case where the victim snapped the defendant in the nose with a 
            towel.}
            \item ``Depraved heart'' killing/implied malice 
            (unintentional--e.g., gross recklessness).
        \end{enumerate}
        \item Both degrees require malice aforethought:
        \begin{enumerate}
            \item Intent to kill.
            \item Intent to cause grievous bodily injury.
            \item Depraved or abandoned heart.
            \item Intent to commit a felony.
        \end{enumerate}
    \end{enumerate}
    \item \textbf{Manslaughter.}
    \begin{enumerate}
        \item Voluntary: requires provocation.
        \item Involuntary: includes negligence and recklessness 
        (unintentional).
    \end{enumerate}
\end{enumerate}
&
\begin{enumerate}
    \item \textbf{Murder.}
    \begin{enumerate}
        \item With purpose.
        \item With knowledge.
        \item With recklessness (unintentional). Recklessness is presumed if 
        the actor is engaged in one of several enumerated felonies (this is 
        the MPC's replacement of the felony-murder rule).
    \end{enumerate}
    \item \textbf{Manslaughter.}
    \begin{enumerate}
        \item With recklessness (unintentional).
        \item Under extreme mental or emotional distress.
    \end{enumerate}
    \item \textbf{Negligent homicide.}
\end{enumerate}
\\ 
\hline
\end{tabular}

\subsection{Intentional Killing}

\subsubsection{Murder}

\begin{enumerate}
    \item Premeditation is usually the distinction between first- and 
    second-degree murder.
    \item \textbf{The MPC does not distinguish between first- and 
    second-degree murder.} Culpability is evaluated at sentencing.
    \item There are four common law definitions of murder:\footnote{Casebook 
    p. 236.}
    \begin{enumerate}
        \item Intent to kill.
        \item Intent to cause grievous bodily harm.
        \item ``Depraved-heart murder'' (i.e., extreme recklessness regarding 
        homicidal risk).
        \item Intent to commit a felony.
    \end{enumerate}
    \item In 1794, Pennsylvania introduced the idea of degrees.
    \item The MPC recognizes three kinds of criminal homicide: murder, 
    manslaughter, and negligent homicide.
    \item Premeditation/deliberation: \textbf{in cold blood}. See 
    \emph{Guthrie} (not enough time for premeditation to occur) and 
    \emph{Midgett}/\emph{Forrest} (proving premeditation from circumstantial 
    evidence) below.
    \item Provocation: \textbf{in hot blood}. See \emph{Girouard} (verbal 
    ll
    provocation) below.
    \item There are \textbf{four elements of the common law provocation 
    defense}, all of which must be present:
    \begin{enumerate}
        \item The defendant acted in the heat of passion.
        \item The passion was the result of provocation.
        \item The defendant did not have a cooling off period.
        \item There is a causal link between the provocation, the passion, and 
        the homicide.
        \item If a \textbf{reasonable person} would have been provoked under 
        the circumstances, the \textbf{provocation mitigates murder to 
        manslaughter.}
    \end{enumerate}
    \item \emph{The MPC does not recognize provocation} as a defense. It uses 
    the \textbf{extreme mental or emotional distress (EMED)} 
    standard.\footnote{MPC \S\ 210.3(1)(b).} Proving EMED is a very low 
    threshold. A jury then determines whether a reasonable person would have 
    been distressed under the circumstances as the person believed them to be. 
    The factfinder may consider the defendant's situation but \emph{not} the 
    defendant's idiosyncratic moral values. See \emph{Casassa} below.
    \item The MPC \textbf{does not require a cooling off period for EMED}.
\end{enumerate}

\paragraph{Premeditation vs. Intent: \emph{State v. Guthrie}}
~\\\\
Premeditation is not the same as intent. Premeditation requires time to become 
fully conscious of what is intended. The minimum time necessary for 
premeditation is not exact, but courts have defined it as \enquote{long enough 
to afford a reasonable man time to subject the nature of his response to a 
\enquote{second look.}}\footnote{\emph{People v. Morrin.}}

\begin{enumerate}
    \item The defendant stabbed and killed a coworker after the coworker 
    taunted him and snapped him in the nose with a towel.
    \item The trial court found him guilty of first-degree murder. The 
    defendant argued that the trial court's instructions to the jury were 
    improper because ``the terms wilful, deliberate, and premeditated were 
    equated with a mere intent to kill.''
    \item The appellate court agreed with the defendant that ``premeditation'' 
    cannot be synonymous with intent---rather, it must be long enough for the 
    defendant to be ``fully conscious of what he intended.'' Reversed.
\end{enumerate}

\paragraph{Proving Premeditation: \emph{Midgett v. State}}
~\\\\
Circumstantial evidence can prove premeditation, but proving it can be 
difficult.

\begin{enumerate}
    \item The defendant repeatedly abused his young son, who died from the 
    injuries. The trial court found him guilty of first-degree murder, which 
    required premeditation and deliberation.
    \item The defendant argued that there was no premeditation, and the 
    Supreme Court of Arkansas agreed.
    \item The dissent argued that symptoms of malnourishment indicated 
    starvation, but the majority argued that the evidence did not prove 
    starvation.
    \item Shortly after this case, the Arkansas legislature amended its 
    criminal code to broaden first-degree murder to include ``extreme 
    indifference to the value of human life'' of people fourteen years old or 
    younger.
\end{enumerate}

\paragraph{Circumstantial Evidence of Premeditation: \emph{State v. Forrest}}
~\\\\
Premeditation can be proved or disproved on the basis of several 
circumstantial factors, including:
\begin{enumerate}
    \item Want of provocation on part of the deceased.
    \item The defendant's conduct and statements (including threats) before, 
    during, and after the killing.
    \item Ill will between the parties.
    \item Dealing of lethal blows after the victim has been rendered helpless.
    \item Evidence of especially brutal killing.
    \item Nature and number of the victim's wounds.
\end{enumerate}

\begin{enumerate}
    \item The defendant shot and killed his terminally ill father in the 
    hospital. The trial court convicted him of first-degree murder.
    \item The defendant argued that there was no premeditation or 
    deliberation, and therefore no evidence to prove first-degree murder.
    \item The appellate court upheld the conviction, noting that premeditation 
    must be proved (or disproved) by circumstantial evidence, including 
    provocation from the victim, the defendant's conduct and statements, ill 
    will between the parties, lethal blows after the victim was rendered 
    helpless, and evidence of an especially brutal killing.
    \item In this case, the court found that the victim was laying helpless 
    and did nothing to provoke the defendant, and that the defendant had 
    earlier made statements about ``putting his father out of his misery.'' It 
    upheld the jury instructions regarding first-degree murder.
\end{enumerate}

\subsubsection{Manslaughter}

\begin{enumerate}
    \item \textbf{Provocation can mitigate murder to manslaughter.} Common law 
    and the MPC diverge on what constitutes provocation.
    \item \textbf{Five elements of common law provocation}, all of which must 
    be present:\footnote{Casebook p. 267.}
    \begin{enumerate}
        \item There must have been adequate provocation.
        \item The killing must have been in the heat of passion.
        \item There must not have been a cooling off period.
        \item There must have been a causal connection between the 
        provocation, the passion, and the fatal act.
        \item (The sorts of provocations that courts have allowed as defenses 
        at common law are the sort of actions that have offended traditional 
        notions of a man's honor---e.g., catching a wife in the act with 
        another man.)
    \end{enumerate}
    \item \textbf{The MPC does not recognize provocation.} It replaces 
    provocation with an \textbf{``emotional disturbance''} test.\footnote{MPC 
    § 210.3(1)(b) at Casebook p. 1000.} Under the MPC, homicide constitutes 
    manslaughter when:
    \begin{enumerate}
        \item The homicide is committed ``under the influence of extreme 
        mental or emotional disturbance.''
        \item There is a reasonable explanation or excuse for the mental or 
        emotional disturbance under the circumstances as the defendant 
        believed them to be. In other words, the excuse is reasonable if a 
        reasonable person in the defendant's situation would have been 
        disturbed.
        \item (The cooling off period is not an issue under the MPC 
        definition.)
    \end{enumerate}
    \item For a gender critique of the MPC's rules of provocation, see 
    Victoria Nourse, ``Passion's Progress.''
    \item \textbf{Words are usually not enough to constitute provocation.} 
    Words can be sufficient if they accompany a threat of intent and ability 
    to cause bodily harm.\footnote{Casebook pp. 267--68.} Some jurisdictions 
    allow \textbf{``informative words''} (e.g, ``your husband is having an 
    affair with ...'') to constitute provocation.
\end{enumerate}

\paragraph{Verbal Provocation: \emph{Girouard v. State}}
~\\\\
Words do not constitute provocation without threat of bodily harm. Some 
jurisdictions, however, have recognized ``informational'' words as sufficient 
provocation (e.g., ``your spouse is sleeping with X'').

\begin{enumerate}
    \item The defendant stabbed and killed his wife after she taunted him 
    relentlessly. The trial court, in a bench trial, convicted him of second 
    degree murder.
    \item The defendant argued on appeal that the rule of provocation should 
    be expanded to include verbal provocation. The appellate court relied on 
    the rule that for provocation to mitigate a charge of murder, it must be 
    ``calculated to inflame the passion of a reasonable man and tend to cause 
    him to act for the moment from passion to reason.'' The standard is 
    objective.
    \item The court found that words can constitute adequate provocation if 
    they accompany intent and ability to cause bodily harm. That was not the 
    case in this scenario, however. The court upheld the second-degree murder 
    conviction.
\end{enumerate}

\paragraph{Emotional Disturbance: \emph{People v. Casassa}}
~\\\\
Emotional disturbance is a defense only if a reasonable person would have been 
disturbed under the circumstances. Emotional disturbance is an objective 
standard.

\begin{enumerate}
    \item The defendant stabbed and killed his neighbor out of jealousy. The 
    trial court found him guilty of second-degree murder.
    \item The defendant argued he was acting under ``extreme emotional 
    disturbance,'' which would reduce the charge to manslaughter.
    \item The appellate court reasoned that the emotional disturbance must 
    meet an objectively reasonable standard. In this case, the disturbance was 
    a result of the defendant's unique mental state---i.e., a reasonable 
    person would not have been so emotionally disturbed under the 
    circumstances. Affirmed.
\end{enumerate}

\subsubsection{Unintentional Killing}

\begin{enumerate}
    \item \textbf{Implied malice} is required to prove murder from an 
    unintentional killing.
    \item At common law, implied malice requires a \textbf{``depraved 
    heart''}---i.e., it involves acting with a conscious disregard for human 
    life and conduct involving a high probability of death. See \emph{Knoller} 
    below.
    \item \textbf{The MPC does not use the depraved heart standard.} Under the 
    MPC, ordinary recklessness proves manslaughter. It proves murder when the 
    actor's ``conscious disregard for the risk, under the circumstances, 
    manifests extreme indifference to the value of human 
    life''\footnote{Casebook p. 303.}---i.e., when the actor behaves with 
    \textbf{gross recklessness}.
    \item At common law, intent to cause grievous bodily injury is sufficient 
    to establish ``malice aforethought.'' The MPC does not adopt this 
    approach---instead, it handles such cases under the standard of extreme 
    recklessness.\footnote{Casebook p. 304.}
    \item Reckless murder and reckless manslaughter are distinguished by the 
    nature of the risk involved.
    \item Reckless manslaughter and negligent homicide are distinguished by 
    the defendant's consciousness of the risk. See \emph{Hernandez} and 
    \emph{Williams} below.
    \item Ordinary (i.e., civil) negligence is generally not sufficient to 
    establish criminal liability, although some states allow it. Generally 
    gross negligence is required.
\end{enumerate}

\paragraph{Implied Malice: \emph{People v. Knoller}}
~\\\\
The Supreme Court of California held that implied malice requires awareness of 
a risk of death, not just serious bodily harm.

\begin{enumerate}
    \item Defendants came into possession of two large, aggressive Presa 
    Canario dogs. They'd been warned repeatedly about the dogs' dangerously 
    aggressive behavior. The dogs killed a woman in the hallway of the 
    defendants' apartment building.
    \item The trial court held that a murder charge required conduct involving 
    ``a high probability of resulting in the death of another.'' The jury 
    found the defendants guilty. The court ordered a new trial.
    \item The appellate court granted defendants' motion for a new trial on 
    the grounds that Knoller did not know that her conduct involved a high 
    probability of death. The appellate court reversed the order for a new 
    trial, holding that the standard for second-degree murder should be 
    ``conscious disregard of the risk of serious bodily injury to another,'' 
    rather than a high probability of death.
    \item The Supreme Court of California focused on the issue of implied 
    malice as an element of murder. It uses two definitions of implied malice: 
    (1) the \emph{Thomas} test: ``wanton disregard for human life, and (2) the 
    \emph{Phillips} test: ``conscious disregard for human life.'' The tests 
    articulate the same standard, but the court prefers the second for 
    clarity.
    \item The Supreme Court reversed the appellate court, holding that 
    \textbf{implied malice requires awareness of a risk of death, not just 
    serious bodily harm}. It also held that the trial court erred in its 
    interpretation of the test. The trial court held the awareness of a high 
    probability of death to be a subjective perception, but it's actually an 
    objective standard. The subjective component is that the defendant must 
    have acted with ``conscious disregard for human life.'' (Not all 
    jurisdictions require conscious disregard to establish implied malice.)
\end{enumerate}

\paragraph{Proving Negligence: \emph{State v. Hernandez}}
~\\\\
To prove negligence, the prosecution must show that the defendant was unaware 
of a substantial and unjustifiable risk. If the defendant was aware of the 
risk, the mental state is recklessness, not negligence.

\begin{enumerate}
    \item The defendant killed a woman while he was driving drunk. The trial 
    court convicted him of involuntary manslaughter. The issue on appeal was 
    whether stickers and pins inside the car with ``drinking slogans''---``The 
    more I drink the better you look,'' etc.---were admissible evidence to help 
    establish the elements of involuntary manslaughter, which the relevant 
    Missouri statute (deriving from the MPC) defined as (1) criminal 
    negligence and (2) resulting death. Criminal negligence is the culpable 
    failure to perceive a substantial and unjustifiable risk.
    \item The government introduced the slogans as evidence that the defendant 
    was aware of the risk of driving drunk. But this was a prosecutorial 
    error, because the statute only criminalizes negligence, and if the 
    defendant had been aware of the risk, he could not have acted negligently 
    (though he may have acted recklessly). Therefore, the evidence was 
    inadmissible in the state's attempt to prove the elements of involuntary 
    manslaughter. (The prosecution probably could have proved second-degree 
    murder based on recklessness.)
    \item The appellate court held that the slogans served only to illustrate 
    the defendant's character. Reputation and character testimony were 
    inadmissible here. Reversed.
    \item The dissent argued that at least three of the slogans indicated that 
    alcohol can impair perception, and that the defendant therefore should 
    have been aware of the substantial risks involved with drinking and 
    driving. It also suggested that the defendant's intoxication might have 
    prevented him from perceiving the risk.
\end{enumerate}

\paragraph{Omissions and Negligence: \emph{State v. Williams}}
~\\\\
Omissions can constitute negligence. The typical standard for negligent 
homicide is gross negligence, not ordinary negligence. Theorists contest 
the justification of punishment for negligent homicide.

\begin{enumerate}
    \item The defendants' infant child died when they failed to seek medical 
    attention for a tooth infection that became a gangrenous abscess.
    \item \textbf{At common law, involuntary manslaughter required gross 
    negligence, not just ordinary negligence.} Washington State law, however, 
    only required ordinary negligence.
    \item The trial court found the defendants guilty of involuntary 
    manslaughter. The appellate court affirmed.
    \item At common law, the defendants likely would not have been convicted 
    of manslaughter.
    \item The appellate court found negligence.
    \item Washington redrafted its code in 1975. Today, ``[c]riminal homicide 
    convictions on the basis of ordinary negligence are nearly non-existent.''
    \item Is it appropriate to punish negligent homicide?
    \begin{enumerate}
        \item The utilitarian argument against punishment is that negligent 
        actors cannot be deterred. The MPC drafters rejected this argument, 
        arguing that the threat of deterrence encourages people to act with 
        greater care.
        \item The retributivist argument is that an actor cannot be morally 
        culpable for actions that he does not know he is taking. Stephen 
        Garvey argues that culpability exists in the failure to exercise 
        self-control ``over desires that influence the formation and awareness 
        of one's beliefs.'' Jerome Hall argues that blame is sometimes 
        appropriate in response to negligence, but punishment is 
        not.\footnote{Casebook pp. 312--13.}
    \end{enumerate}
\end{enumerate}

\subsubsection{Felony-Murder}

\begin{enumerate}
    \item \textbf{Felony-murder rule}: ``...one is guilty of murder if a death results from conduct during 
    the commission or attempted commission of any felony.''\footnote{ALI 
    Commentary to MPC \S\ 210.2, from \emph{Understanding Criminal Law} p. 
    510.} Rationales:
    \begin{enumerate}
        \item Deterrence: encourage criminals to commit felonies more safely.
        \item Transferred intent: intent to commit the felony implies intent 
        to commit the homicide.
        \item Retribution: the defendant's culpability for committing the 
        felony demonstrates an evil mind.
        \item General culpability.
    \end{enumerate}
    \item The rule can typically only be invoked for \textbf{inherently dangerous 
    felonies}.
    \item \textbf{Enumerated felonies}: if the felony is enumerated in 
    statute, the homicide is first-degree murder. All others are 
    second-degree.
    \item \textbf{The MPC does not identify felony-murder as a separate 
    offense.} Instead, offenses that would have triggered the felony-murder 
    rule at common law would likely fall under reckless murder under the MPC. 
    The proliferation of statutory (\emph{malum prohibitem}) felonies led the 
    MPC to \textbf{enumerate the specific felonies that can constitute the bases for 
    reckless murder}.\footnote{The enumerated felonies in the MPC are ``robbery, 
    rape or deviate sexual intercourse by force or threat of force, arson, 
    burglary, kidnapping or felonious escape.'' MPC \S\ 210.2(1)(b), Casebook 
    p. 1000.}
    \item \textbf{Independent felony (or merger) limitation}: the felony 
    murder rule only applies if the felony is independent of the homicide. 
    Assaultive felonies (those that are ``on the road'' to homicide) do not 
    trigger the felony-murder rule. Rationales:
    \begin{enumerate}
        \item The limitation prevents every felonious assault that results in 
        homicide from being punished as second-degree murder.
        \item If A negligently kills B, A would be guilty of negligent 
        homicide. Negligent homicide is itself a felony. If the felony-murder 
        rule applied, A would be guilty of murder, and \textbf{the crime of 
        negligent homicide would evaporate}. The same is true for voluntary 
        manslaughter (e.g., with provocation).
        \item If the predicate felony is assaultive, \textbf{it cannot 
        logically be done more safely}.
    \end{enumerate}
    \item \textbf{\emph{Res gestae} requirement}: the felony-murder rule only 
    applies when a killing occurs during the commission or attempted 
    commission of a felony (which includes any proximate causal relationship).
    \item You cannot attempt to commit felony murder. % todo why?
    \item At common law, any felony was punishable by death.
    \item At common law, any homicide committed while committing (or 
    attempting to commit) one of several enumerated felonies was considered 
    murder. The homicide was a strict liability offense.
    \item Most felony murders could also be charged as depraved heard murder. 
    The felony-murder rule is easier for prosecutors because it allows them to 
    prove murder without establishing mens rea. Some argue that it establishes 
    strict liability for homicides that occur during the commission of a 
    felony, while others hold that committing a felony constitutes implied 
    malice.
\end{enumerate}

\paragraph{Applying the Felony-Murder Rule: \emph{People v. Fuller}} 
~\\\\
The felony-murder rule is harsh.

\begin{enumerate}
    \item The defendant had been breaking into cars in a parking lot when the 
    police noticed him, and a chase ensued. He accidentally killed a driver 
    while involved in a high speed car chase.
    \item The Court of Appeal ruled that the trial court had erred in striking 
    the first-degree murder count. The appellate court allowed the prosecution 
    for first-degree murder under the felony-murder rule---but it noted that 
    if it were ``starting from a clean slate,'' it would not allow the 
    prosecution because the original felony, burglary, was not dangerous to 
    human life.
\end{enumerate}

\paragraph{Roth and Sundby, ``The Felony-Murder Rule''}

\begin{enumerate}
    \item The US is the only western country that recognizes the felony-murder 
    rule.
    \item The rule is meant to (1) deter accidental killings during felonies 
    as well as (2) the felonies themselves. On (1), how can you deter an 
    unintentional act? On (2), there is doubt that stricter punishments deter 
    serious crimes, and it makes more sense to punish the intended conduct 
    (e.g., carrying a deadly weapon) rather than the unintended killing. It 
    may also creative perverse incentives to commit homicide during the 
    commission of a felony.
    \item Transferred intent is not a valid justification for the felony 
    murder rule because of the differences in mens rea for the felony 
    and for murder.
    \item Justifying a murder charge on a retributivist justification is a 
    regression to the primitive ``evil mind'' theory of common law, and it 
    defies the principle of proportional punishment.
\end{enumerate}

\paragraph{Crump and Crump, ``In Defense of the Felony-Murder Doctrine''}

\begin{enumerate}
    \item ``Felony-murder reflects a social judgment'' that felonies involving 
    killing are more serious than non-lethal felonies.
    \item The rule distinguishes crimes that cause death, thereby 
    ``reinforcing the reverence for human life.''
    \item Core disagreement with Roth and Sundby: punishing negligent killings 
    \emph{can} deter future negligence. Also, felons who killed intentionally 
    might testify that the killings were accidental; the felony-murder rule 
    denies them this defense.
    \item A clear felony-murder rule is less confusing to juries, so it leads 
    to more consistent results. Also, by simplifying the questions involved, 
    it makes administration more efficient.
\end{enumerate}

\paragraph{Tomkovicz, ``The Endurance of the Felony-Murder Rule''}

\begin{enumerate}
    \item Restricting the felony-murder rule to certain types of felonies 
    enhances its fairness, helping the doctrine survive.
\end{enumerate}

\paragraph{Inherently Dangerous Felonies: \emph{People v. Howard}}
~\\\\
The felony-murder rule is limited to inherently dangerous felonies.

\begin{enumerate}
    \item The defendant was driving a stolen car without a rear license plate. 
    A chase ensued when police tried to pull him over. During the chase, the 
    defendant hit and killed another driver.
    \item The trial jury convicted the defendant of second-degree murder. The 
    appellate court affirmed, rejecting the defendant's claim that he could 
    not be charged with second-degree murder because of California precedent 
    rejecting the felony-murder rule for felonies that are not inherently 
    dangerous.
    \item The California Supreme Court looked at the statute defining 
    high-speed chases.\footnote{\S\ 2800.2, Casebook p. 329.} It noted 
    that in 1996, the legislature significantly broadened the statutory 
    definition of ``willful or wanton disregard for the safety of persons or 
    property'' to include any flight from a police officer involving three 
    traffic violations. It concluded that a violation of this statute could 
    result from violating a few minor traffic rules, and therefore any 
    violating ``is not, in the abstract, inherently dangerous to human life.'' 
    Therefore, the prosecution could not rely on the felony-murder rule. 
    Reversed.
    \item Brown, concurring and dissenting: this interpretation of the statute 
    defies common sense. The conviction should be overturned, but only because 
    the felony-murder rule should be removed entirely.
    \item Baxter, dissenting: ``there is no doubt that the defendant committed 
    exactly the reckless endangerment of human life forbidden by the 
    statute.''
\end{enumerate}

\paragraph{Independent Felony (or Merger) Limitation: \emph{People v. Smith}}
~\\\\
If the homicide occurred ``on the road'' to the predicate felony, the felony 
murder rule does not apply.

\begin{enumerate}
    \item Defendant was abusing her child, who accidentally fell, hit her 
    head, and died of respiratory arrest. The predicate felony was child 
    abuse. The trial court applied the felony 
    murder rule to convict her of second-degree murder.
    \item \emph{People v. Ireland}: ``We therefore hold that a second-degree 
    felony-murder instruction may not properly be given when it is based upon 
    a felony which is an integral part of the homicide and which the evidence 
    produced by the prosecution shows to be an offense included in fact within 
    the offense charged.''\footnote{Casebook p. 335.}
    \item \emph{People v. Wilson}: The felony-murder rule cannot be applied to 
    cases where the action would not be felonious but for the assault, and the 
    assault is an integral part of the homicide.
    \item \emph{People v. Sears}: If assault is intended against one person 
    but results in the accidental killing of another, the felony-murder rule 
    should not apply (because it would carry harsher punishments than if the 
    intended victim was killed, in which case the felony-murder rule would not 
    not apply).
    \item \emph{People v. Burton}: The felony-murder rule can apply if the 
    underlying violent action was committed with an ``independent felonious 
    purpose.'' For instance, in an armed robbery case where an accidental 
    killing results, the rule applies because the underlying purpose was to 
    rob, not to assault.
    \item The felony-murder rule does not apply in this case. Reversed.
\end{enumerate}

\paragraph{Liability for a Non-Felon's Actions: \emph{State v. Sophophone}}
~\\\\
The felony-murder rule does not apply when a law enforcement officer causes 
the death.

\begin{enumerate}
    \item The defendant and three accomplices broke into a house. The police 
    arrived and shot one of the accomplices. The defendant was charged, among 
    other things, with felony-murder. The trial court convicted him on all 
    counts.
    \item Under Kansas state law, aggravated burglary counted as one of the 
    inherently dangerous felonies that triggers the felony-murder rule.
    \item The defendant argued that he was in custody at the time of his 
    accomplice's death and cannot therefore be held liable.
    \item The two issues in question were (1) ``whether the felony-murder rule 
    should apply when the fatal act is performed by a non-felon'' and (2) when 
    the felony stopped. There are two approaches:
    \begin{enumerate}
        \item \textbf{Agency approach} (the majority rule): The felony-murder 
        rule does not apply when the person who causes the death is a 
        non-felon. The killing was the result of actions contrary to the 
        intentions of the felon.
        \item \textbf{Proximate causation approach}: The felony-murder rule 
        applies. A felon is responsible for the consequences of the actions he 
        sets in motion.
    \end{enumerate}
    \item \textbf{\emph{Res gestae}} (``things done''): The felony-murder rule 
    applies when a killing occurs during the commission (or attempted 
    commission) of a felony. Most courts also apply it in the aftermath, e.g., 
    during a getaway.
    \item There must also be causal relationship between the death and the 
    felony. The cause must be proximate.
    \item The court holds that the rule does not apply. ``...we believe that 
    making one criminally responsible for the lawful acts of a law enforcement 
    officer is not the intent of the felony-murder statute as it is currently 
    written.''\footnote{Casebook p. 340.}
    \item Dissent: nothing in the statute requires the court to adopt the 
    ``agency'' approach. ``This set of events could have very easily resulted 
    in the death of a law enforcement officer, and in my opinion this is 
    exactly the type of case the legislature had in mind when it adopted the 
    felony-murder rule.''\footnote{Casebook p. 341.}
\end{enumerate}

\paragraph{Misdemeanor Manslaughter}

\begin{enumerate}
    \item ``An unintended homicide that occurs during the commission of an 
    unlawful act not amounting to a felony constitutes common law involuntary 
    manslaughter.''\footnote{Casebook p. 343}
    \item Manslaughter convictions have been upheld in cases where the act is 
    morally culpable but not technically criminal---e.g., someone attempted to 
    commit suicide with a gun, someone attempted to intervene, and the one who 
    intervened was accidentally shot and killed.
    \item Not all jurisdictions apply the misdemeanor manslaughter rule, and 
    some only apply it for \emph{malum in se} misdemeanors.
    \item The MPC does not follow the misdemeanor manslaughter rule because 
    it's too harsh.
\end{enumerate}
