\section{Overview}

\begin{enumerate}
    \item Basic principles of criminal law.
    \begin{enumerate}
        \item Legality principle, reasonable doubt, retributivism, 
        utilitarianism, sentencing, proportionality, institutional competency, 
        statutory clarity, lenity.
    \end{enumerate}
    \item Elements of a crime.
    \begin{enumerate}
        \item Actus reus: harm principle, conduct vs. status, omissions, duty, 
        constitutional limitations.
        \item Mens rea: culpability vs. elemental, specific intent vs. general 
        intent, willful blindness, transferred intent, strict liability, 
        public welfare offenses, mistake of fact, mistake of law, MPC 
        subjectivity.
        \item Causation: cause-in-fact, but-for, substantial factor, proximate 
        cause, intervening acts.
    \end{enumerate}
    \item Homicide.
    \begin{enumerate}
        \item Common law vs. MPC classifications, provocation (common law vs. 
        MPC), implied malice/depraved heart, negligence, felony-murder rule, 
        merger limitation, res gestae, misdemeanor manslaughter.
    \end{enumerate}
    \item Rape.
    \begin{enumerate}
        \item Forcible rape: common law definition, forcible resistance 
        requirement, threats, spouse exemption, apprehension of fear, 
        acquaintance rape.
        \item Mens rea: mistake of fact.
        \item Statutory rape: policy goals, reforms, equal protection.
    \end{enumerate}
    \item Defenses.
    \begin{enumerate}
        \item Failure of proof, offense modification, justification, excuse, 
        noexculpatory public policy defense.
        \item Justification: triggering conditions, necessity, 
        proportionality.
        \begin{enumerate}
            \item Self defense: common law vs. MPC, castle doctrine, 
            immediacy vs. imminence, imperfect self-defense, abused spouse 
            syndrome.
            \item Necessity: MPC vs. common law, emergencies, civil 
            disobedience
        \end{enumerate}
        \item Excuse: theories (utilitarian, causation, character, free 
        choice).
        \begin{enumerate}
            \item Duress: coercion, immediacy, homicide.
            \item Intoxication: voluntary vs. involuntary, specific intent 
            crimes.
            \item Insanity: deific decree, tests (right-wrong, M'Naghten, 
            irresistible impulse, Durham/product, MPC/ALI).
            \item Diminished capacity: continuum of competence, mens rea 
            variant vs. partial responsibility variant.
            \item Infancy: under seven, seven to fourteen, above fourteen.
        \end{enumerate}
    \end{enumerate}
    \item Inchoate offenses.
    \begin{enumerate}
        \item Rationales for punishment, law enforcement discretion.
        \item Attempt: incomplete vs. complete, merger, conduct vs. results 
        crimes, common law tests (first step, last step, physical proximity, 
        dangerous proximity, indispensable element, probable desistance, 
        abnormal step, res ipsa loquitur/unequivocality, legal and factual 
        impossibility, abandonment.
        \item Assault: original definition, current status.
        \item Solicitation: merger, abandonment, incomplete communication.
        \item Conspiracy: inference, merger, Pinkerton liability, overt act, 
        vicarious liability (MPC), prosecutors' love, multiple crimes, wheel 
        vs. chain, withdrawal, inferring intent, Wharton's Rule, legislative 
        exemption defense.
    \end{enumerate}
    \item Accomplice liability.
    \begin{enumerate}
        \item Derivative liability, conduct, common law categories vs. modern 
        status, justification vs. excuse, legislative exemption defense, 
        abandonment, innocent agency doctrine.
    \end{enumerate}
\end{enumerate}
